---
title: "EHR Data Quality Control and Analysis"
author: "Fereshteh Izadi"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)

## Load Data

```{r load-data}
# Set path to data directory
data_path <- "C:/Users/User/Downloads/test_data/data/dest/"

# Load main datasets
patients <- read_csv(paste0(data_path, "patients.csv"))
conditions <- read_csv(paste0(data_path, "conditions.csv"))
observations <- read_csv(paste0(data_path, "observations.csv"))
medications <- read_csv(paste0(data_path, "medications.csv"))
encounters <- read_csv(paste0(data_path, "encounters.csv"))

# Load dictionaries
dict_snomed <- read_csv(paste0(data_path, "dictionary_snomed.csv"))
dict_rxnorm <- read_csv(paste0(data_path, "dictionary_rxnorm.csv"))
dict_loinc  <- read_csv(paste0(data_path, "dictionary_loinc.csv"))
```

## Patients Quality Control (QC)

```{r patients-qc}
# Convert date columns
patients$BIRTHDATE <- as.Date(patients$BIRTHDATE)
patients$DEATHDATE <- as.Date(patients$DEATHDATE)

# Calculate age
patients$AGE <- as.numeric(floor(interval(patients$BIRTHDATE, Sys.Date()) / years(1)))

# Filter out invalid data
patients <- patients %>%
  filter(AGE >= 0 & AGE <= 100) %>%
  filter(is.na(DEATHDATE) | BIRTHDATE <= DEATHDATE) %>%
  filter(!(LAT == 0 & LON == 0)) %>%
  filter(LAT >= -90 & LAT <= 90) %>%
  filter(LON >= -180 & LON <= 180) %>%
  filter(!(HEALTHCARE_EXPENSES == 0 & HEALTHCARE_COVERAGE > 5000)) %>%
  filter(GENDER %in% c("M", "F", "", NA)) %>%
  filter(!((PREFIX == "Mr." & GENDER == "F") | 
           (PREFIX %in% c("Mrs.", "Ms.", "Miss") & GENDER == "M")))

# Remove US state + UK city mismatches
us_states <- state.abb
uk_cities <- c("London", "Manchester", "Leeds", "Birmingham", "Glasgow", "Liverpool")
patients <- patients[!(patients$CITY %in% uk_cities & patients$STATE %in% us_states), ]

# Fix race coding
patients$RACE[patients$RACE == "XJniDSe"] <- "other (possibly miscoded)"

# ZIP formatting
patients$ZIP <- sprintf("%05d", patients$ZIP)
```

## Encounters QC

```{r encounters-qc}
encounters$START <- as.POSIXct(encounters$START)
encounters$STOP <- as.POSIXct(encounters$STOP)

encounters <- encounters %>%
  filter(is.na(STOP) | STOP >= START) %>%
  filter(BASE_ENCOUNTER_COST >= 0, TOTAL_CLAIM_COST >= 0, PAYER_COVERAGE >= 0) %>%
  filter(PATIENT %in% patients$Id)
```

## Conditions QC

```{r conditions-qc}
conditions$START <- as.Date(conditions$START)
conditions$STOP <- as.Date(conditions$STOP)

conditions <- conditions %>%
  filter(is.na(STOP) | START <= STOP) %>%
  filter(PATIENT %in% patients$Id) %>%
  filter(ENCOUNTER %in% encounters$Id) %>%
  filter(CODE %in% dict_snomed$CODE)

conditions <- left_join(conditions, dict_snomed, by = "CODE")

# Add 0/1 flags for socioeconomic factors based on DESCRIPTION content
conditions <- conditions %>%
  mutate(
    factor_education = as.integer(grepl("education|high school|primary school|higher education", DESCRIPTION, ignore.case = TRUE)),
    factor_employment = as.integer(grepl("employment|unemployed|labor force", DESCRIPTION, ignore.case = TRUE)),
    factor_stress = as.integer(grepl("stress|anxiety|panic", DESCRIPTION, ignore.case = TRUE)),
    factor_refugee = as.integer(grepl("refugee", DESCRIPTION, ignore.case = TRUE)),
    factor_criminal = as.integer(grepl("criminal", DESCRIPTION, ignore.case = TRUE)),
    factor_violence = as.integer(grepl("violence|abuse|victim", DESCRIPTION, ignore.case = TRUE)),
    factor_transport = as.integer(grepl("transport", DESCRIPTION, ignore.case = TRUE)),
    factor_drugs_alcohol = as.integer(grepl("alcohol|misuses drugs", DESCRIPTION, ignore.case = TRUE)),
    factor_social_isolation = as.integer(grepl("social|isolation|limited contact", DESCRIPTION, ignore.case = TRUE)),
    factor_military = as.integer(grepl("armed forces|military", DESCRIPTION, ignore.case = TRUE)),
    factor_housing = as.integer(grepl("housing", DESCRIPTION, ignore.case = TRUE))
  )


```

## Observations QC

```{r observations-qc}
observations$DATE <- as.POSIXct(observations$DATE)
observations$CODE <- as.character(observations$CODE)
dict_loinc$CODE <- as.character(dict_loinc$CODE)

observations <- observations %>%
  left_join(dict_loinc, by = "CODE") %>%
  filter(!is.na(DESCRIPTION), !is.na(VALUE), VALUE != "", !is.na(TYPE), TYPE != "")

observations$VALUE_NUMERIC <- suppressWarnings(as.numeric(observations$VALUE))
observations <- observations[observations$TYPE != "numeric" | !is.na(observations$VALUE_NUMERIC), ]

# Weight-kg mismatch
mismatch <- grepl("weight", tolower(observations$DESCRIPTION)) & !grepl("kg", tolower(observations$UNITS))
observations <- observations[!mismatch, ]

observations <- observations %>%
  filter(PATIENT %in% patients$Id, ENCOUNTER %in% encounters$Id)
```

## Medications QC

```{r medications-qc}
medications$START <- as.POSIXct(medications$START)
medications$STOP <- as.POSIXct(medications$STOP)

medications <- medications %>%
  filter(is.na(STOP) | START <= STOP, PATIENT %in% patients$Id, ENCOUNTER %in% encounters$Id) %>%
  filter(BASE_COST >= 0, PAYER_COVERAGE >= 0, TOTALCOST >= 0)

medications <- medications %>%
  left_join(dict_rxnorm, by = "CODE") %>%
  filter(!is.na(DESCRIPTION)) %>%
  filter(BASE_COST <= TOTALCOST) %>%
  filter(abs((BASE_COST + PAYER_COVERAGE) - TOTALCOST) <= 1)
```

## Descriptive Summaries and Visualizations

```{r summary-plots}
# Gender distribution
ggplot(patients, aes(x = GENDER)) + geom_bar(fill = "steelblue") + theme_minimal()

# Age distribution
ggplot(patients, aes(x = AGE)) + geom_histogram(binwidth = 5, fill = "darkorange") + theme_minimal()

# Ethnicity and Race
ggplot(patients, aes(x = ETHNICITY)) + geom_bar(fill = "purple") + coord_flip() + theme_minimal()
ggplot(patients, aes(x = RACE)) + geom_bar(fill = "green") + coord_flip() + theme_minimal()
```

## Top Medications, Conditions, and Labs

```{r top-summary}
top_meds <- medications %>%
  group_by(PATIENT, DESCRIPTION) %>% summarise(n = n()) %>%
  group_by(DESCRIPTION) %>% summarise(total = sum(n)) %>% arrange(desc(total)) %>% head(10)

conditions_summary <- conditions %>%
  group_by(DESCRIPTION) %>% summarise(n = n()) %>% arrange(desc(n))
top_conditions <- head(conditions_summary, 10)
least_conditions <- tail(conditions_summary, 10)

top_labs <- observations %>%
  group_by(DESCRIPTION) %>% summarise(n = n()) %>% arrange(desc(n)) %>% head(10)
```

## Medications by Socioeconomic Group

### Medications for Unemployed Patients

```{r unemployed_meds}

# Step 1: Find encounters where unemployment is recorded
unemployed_encounters <- conditions %>%
  filter(grepl("unemployed|not in labor force", DESCRIPTION, ignore.case = TRUE)) %>%
  distinct(PATIENT, ENCOUNTER)

# Step 2: Join to medications
medications_unemployed <- medications %>%
  semi_join(unemployed_encounters, by = c("PATIENT", "ENCOUNTER"))

# Step 3: Plot top medications
medications_unemployed %>%
  count(DESCRIPTION, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(DESCRIPTION, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top Medications Among Unemployed Patients",
    x = "Medication", y = "Count"
  ) +
  theme_minimal()
```




## Conditions with Highest Out-of-Pocket Costs

```{r cost-analysis}
cond_cost <- merge(conditions, encounters[, c("Id", "TOTAL_CLAIM_COST", "PAYER_COVERAGE")],
                   by.x = "ENCOUNTER", by.y = "Id") %>%
  group_by(DESCRIPTION) %>%
  summarise(
    avg_claim_cost = mean(TOTAL_CLAIM_COST, na.rm = TRUE),
    avg_patient_pay = mean(TOTAL_CLAIM_COST - PAYER_COVERAGE, na.rm = TRUE)) %>%
  arrange(desc(avg_patient_pay))
```

## Reasons for Visit

```{r reasons}
reasons <- encounters %>%
  filter(!is.na(REASONCODE)) %>%
  count(REASONCODE, sort = TRUE) %>%
  left_join(dict_snomed, by = c("REASONCODE" = "CODE")) %>%
  rename(ReasonDescription = DESCRIPTION)
head(reasons, 20)
```

## Done

```{r final-checks}
summary(patients)
summary(encounters)
summary(medications)
summary(conditions)
summary(observations)
```
